---
title: "Difference-in-Differences I: Classical DiD"
subtitle: "Lecture 3 - Introduction to Causal Inference"
author: "Kevin Li"
format:
  beamer:
    slide-number: true
header-includes: |
  \setbeamertemplate{footline}[frame number]  % Forces slide numbers in footline
  \setbeamertemplate{navigation symbols}{}   % Optional: Removes navigation symbols
editor: visual
---

## New Jersey Employment

New Jersey implemented a minimum wage increase in April 1992. We want to find the causal effect of this on employment in fast food restaurants.

We have average data of all restaurant's employment rate from before and after the change.

```{r}
#| fig-height: 4
#| warning: false
#| message: false
#| fig-align: center

library(tidyverse)
df.nj <- data.frame(
  time = c(0, 1),
  nj = c(20, 24)
)

df.pa <- data.frame(
  time = c(0, 1),
  pa = c(18, 19)
)

df.pot <- data.frame(
  time = c(0, 1),
  nj.pot = c(20, 21)
)

plot1 <- ggplot() +
  geom_point(aes(x = time, y = nj, color = "col2"), data = df.nj) +
  geom_line(aes(x = time, y = nj, color = "col2"), data = df.nj) +
  scale_color_manual(
    name = "Legend",
    values = c("col2" = "darkgreen"),
    labels = c("New Jersey")
  ) +
  theme_bw() +
  scale_x_continuous(breaks = c(0, 1), labels = c("Mar 1992","Dec 1992")) +
  xlab(NULL) + ylab("Employment Rate (Y)") +
  ylim(17, 27) + 
  theme(
    axis.title.x = element_text(size = 14),  # Increase x-axis title size
    axis.title.y = element_text(size = 18),  # Increase y-axis title size
    axis.text.x = element_text(size = 16),   # Increase x-axis tick label size
    axis.text.y = element_text(size = 16),  # Increase y-axis tick label size
    legend.title = element_text(size = 16),  # Increase legend title size
    legend.text = element_text(size = 14)    # Increase legend text (labels) size
  )

plot1
```

## Our Issue

Is our causal effect the Dec 1992 - Mar 1992? No!

Why? Something could have happened between March and December, like a recession or something else.

-   How do we know if those factors are explaining the change in employment, or our treatment?

We can also think of this issue in terms of potential outcomes. Our causal effect in December 1992 is:

$$
\text{Average }\tau_\text{Dec} = \overline Y_\text{Dec}(1) - \overline Y_\text{Dec}(0)
$$

But we do not observe $\overline Y_\text{Dec}(0)$ since New Jersey is treated in Dec.

## Solution: Trend in Control Group

What if we looked at another state who did not have a minimum wage increase during this time, such as Pennsylvania?

Pennsylvania tells us the **trend** in employment $Y$ if no treatment occurred.

```{r}
#| fig-height: 4
#| warning: false
#| message: false
#| fig-align: center

ggplot() +
  geom_point(aes(x = time, y = nj, color = "col2"), data = df.nj) +
  geom_line(aes(x = time, y = nj, color = "col2"), data = df.nj) +
  geom_point(aes(x = time, y = pa, color = "col1"), data = df.pa) +
  geom_line(aes(x = time, y = pa, color = "col1"), data = df.pa) +
  scale_color_manual(
    name = "Legend",
    values = c("col1" = "red",
               "col2" = "darkgreen"),
    labels = c("Pennsylvania", "New Jersey")
  ) +
  theme_bw() +
  scale_x_continuous(breaks = c(0, 1), labels = c("Mar 1992","Dec 1992")) +
  xlab(NULL) + ylab("Employment Rate") +
  ylim(17, 27) +
  theme(
    axis.title.x = element_text(size = 14),  # Increase x-axis title size
    axis.title.y = element_text(size = 18),  # Increase y-axis title size
    axis.text.x = element_text(size = 16),   # Increase x-axis tick label size
    axis.text.y = element_text(size = 16),  # Increase y-axis tick label size
    legend.title = element_text(size = 16),  # Increase legend title size
    legend.text = element_text(size = 14)    # Increase legend text (labels) size
  )
```

## Solution: Parallel Trends

Then, we assume that if New Jersey had hypothetically not received a minimum wage increase, it would have followed the same trend as Pennsylvania.

-   This allows us to estimate New Jersey's counterfactual, and find New Jersey's causal effect in December.

```{r}
#| fig-height: 4
#| warning: false
#| message: false
#| fig-align: center

df.att <- data.frame(
  time = c(1, 1),
  att = c(23.9, 21.1)
)
plot2 <- ggplot() +
  geom_point(aes(x = time, y = nj.pot, color = "col3"), data = df.pot) +
  geom_line(aes(x = time, y = nj.pot, color = "col3"), data = df.pot, linetype = 2) +
  geom_point(aes(x = time, y = nj, color = "col2"), data = df.nj) +
  geom_line(aes(x = time, y = nj, color = "col2"), data = df.nj) +
  geom_point(aes(x = time, y = pa, color = "col1"), data = df.pa) +
  geom_line(aes(x = time, y = pa, color = "col1"), data = df.pa) +
  geom_line(aes(x = time, y = att, color = "col4"), data = df.att, arrow = arrow(length = unit(0.2, "cm"), ends = "both", type = "closed"), linetype = 3) +
  scale_color_manual(
    name = "Legend",
    values = c("col1" = "red",
               "col2" = "darkgreen",
               "col3" = "purple",
               "col4" = "black"),
    labels = c("Pennsylvania", "New Jersey", "New Jersey (Counterfactual)", "Treatment Effect")
  ) +
  theme_bw() +
  scale_x_continuous(breaks = c(0, 1), labels = c("Mar 1992","Dec 1992")) +
  xlab(NULL) + ylab("Employment Rate") +
  ylim(17, 27) +
  theme(
    axis.title.x = element_text(size = 14),  # Increase x-axis title size
    axis.title.y = element_text(size = 18),  # Increase y-axis title size
    axis.text.x = element_text(size = 16),   # Increase x-axis tick label size
    axis.text.y = element_text(size = 16),  # Increase y-axis tick label size
    legend.title = element_text(size = 16),  # Increase legend title size
    legend.text = element_text(size = 14)    # Increase legend text (labels) size
  )

plot2
```

## Classical DiD

The example we just showed is an example of classical difference-in-differences. We have 2 groups of observations:

-   A treated group $G= 1$. These are units that will receive treatment sometime in our study (like New Jersey).

-   An untreated group $G = 0$. These are units that will not receive treatment (like Pennsylvania).

We have 2 time periods.

-   In $t=0$ (pre-treatment), both groups are untreated.

-   In $t=1$ (post-treatment), only the treated group is treated, while the control group remains untreated.

## Estimating Causal Effects

The trend in the untreated group $G=0$ between time periods is:

$$
\text{trend} = \overline Y^{(G=0)}_{t=1} - \overline Y^{(G=0)}_{t=0}
$$

The estimated counterfactual for the treated group is thus:

$$
\overline Y^{(G=1)}_{t=0} + (\text{trend}) \ =\ \overline Y^{(G=1)}_{t=0} + \left(\overline Y^{(G=0)}_{t=1} - \overline Y^{(G=0)}_{t=0}\right)
$$

The estimated causal effect is thus:

$$
\tau_\text{ATT} = \overline Y^{(G=1)}_{t=1} - \underbrace{\left [\overline Y^{(G=1)}_{t=0} + \left(\overline Y^{(G=0)}_{t=1} - \overline Y^{(G=0)}_{t=0}\right) \right]}_\text{estimated counterfactual}
$$

## Why is it Called Difference-in-Differences?

Distribute out the negative sign, and you get:

$$
\tau_\text{ATT} = \underbrace{\left(\overline Y^{(G=1)}_{t=1} - \overline Y^{(G=1)}_{t=0}\right)}_{\Delta \text{ for } G=1} - \underbrace{\left(\overline Y^{(G=0)}_{t=1} - \overline Y^{(G=0)}_{t=0}\right)}_{\Delta \text{ for } G=0} 
$$

-   The first part is the difference in average $Y$ between time periods for the treated group $G=1$.

-   The second part is the difference in average $Y$ between time periods for the control group $G=0$.

The total causal effect is the difference of these two differences. Hence, difference-in-differences.

## Regression Estimator

We can also use a regression estimator, which makes calculating standard errors simpler:

$$
\hat Y_{it} = \hat\alpha +  G_i\hat \gamma +  T_{it}\hat\delta + G_iT_{it} \hat\tau_\text{ATT}
$$

-   $G_{i}$ is the variable indicating the group for the unit (treated, control)

-   $T_{it}$ is the variable indicating the time period for the observation.

-   Standard errors should be clustered at the group $G_i$ level.

(You can confirm this is equal to the previous DiD estimator - plug in the correct values for $G_i$ and $T_{it}$ for each $\overline Y^{(G=G_i)}_{t=T_{it}}$ in the previous equation for DiD).

## Parallel Trends Assumption

DiD depends on the parallel trends assumption.

-   That if the treated group had really been untreated, it would have followed the same trend as the control group.
-   Ex: New Jersey, if it had no minimum wage rise, would have followed the same trend in employment as Pennsylvania.

```{r}
#| fig-height: 4
#| warning: false
#| message: false
#| fig-align: center

plot2
```

## Example of Violating Parallel Trends

Here, we can see the true counterfactual of New Jersey is not the same trend as Pennsylvania. So if we assumed the parallel trends assumption, our causal estimate would be very wrong.

```{r}
#| fig-height: 4
#| warning: false
#| message: false
#| fig-align: center

df.pt1 <- data.frame(
  time = c(0, 1),
  nj.pot = c(20, 16)
)

df.att <- data.frame(
  time = c(1, 1),
  att = c(23.9, 16.1)
)

ggplot() +
  geom_point(aes(x = time, y = nj.pot, color = "col3"), data = df.pot) +
  geom_line(aes(x = time, y = nj.pot, color = "col3"), data = df.pot, linetype = 2) +
  geom_point(aes(x = time, y = nj.pot, color = "col4"), data = df.pt1) +
  geom_line(aes(x = time, y = nj.pot, color = "col4"), data = df.pt1) +
  geom_point(aes(x = time, y = nj, color = "col2"), data = df.nj) +
  geom_line(aes(x = time, y = nj, color = "col2"), data = df.nj) +
  geom_point(aes(x = time, y = pa, color = "col1"), data = df.pa) +
  geom_line(aes(x = time, y = pa, color = "col1"), data = df.pa) +
  geom_line(aes(x = time, y = att, color = "col5"), data = df.att, arrow = arrow(length = unit(0.2, "cm"), ends = "both", type = "closed"), linetype = 3) +
  scale_color_manual(
    name = "Legend",
    values = c("col1" = "red",
               "col2" = "darkgreen",
               "col3" = "purple",
               "col4" = "purple",
               "col5" = "black"),
    labels = c("Pennsylvania", "New Jersey", "New Jersey (Parallel Trends)", "New Jersey (Real Counterfactual)", "Real Treatment Effect")
  ) +
  theme_bw() +
  scale_x_continuous(breaks = c(0, 1), labels = c("Mar 1992","Dec 1992")) +
  xlab(NULL) + ylab("Employment Rate") +
  ylim(15, 27) +
  theme(
    axis.title.x = element_text(size = 14),  # Increase x-axis title size
    axis.title.y = element_text(size = 18),  # Increase y-axis title size
    axis.text.x = element_text(size = 16),   # Increase x-axis tick label size
    axis.text.y = element_text(size = 16),  # Increase y-axis tick label size
    legend.title = element_text(size = 16),  # Increase legend title size
    legend.text = element_text(size = 14)    # Increase legend text (labels) size
  )
```
