---
title: "Difference-in-Differences II: Generalised DiD"
subtitle: "Lecture 4 - Introduction to Causal Inference"
author: "Kevin Li"
format:
  beamer:
    slide-number: true
header-includes: |
  \setbeamertemplate{footline}[frame number]  % Forces slide numbers in footline
  \setbeamertemplate{navigation symbols}{}   % Optional: Removes navigation symbols
editor: visual
---

## More Time Periods

In Classical DiD, we have two time periods for units.

-    $t = -1$ (pre-treatment) and $t=0$ (post-treatment).

In Generalised DiD, we allow for more time periods.

-   Multiple Pre-treatment periods $t = -1, -2, -3, \dots$.

-   Multiple Post-treatment periods $t = 0, 1, 2, 3, \dots$.

For now, we assume panel data - i.e. the observed individuals/units in our data are observed throughout each time period.

-   Ex. individual $A$ has a value for each time period $\dots, -2, -1, 0, 1, 2, \dots$.

## Relative Time Periods

In our data, treatment might start in any time-period, for example, the year $t=2000$.

It is standard to create a new variable called **relative time** $R$ - that indicates the time relative to initial treatment adoption for each cohort.

-   The first treatment year in the relative time variable is always $R=0$.

-   Pre-treatment periods are always negative $R ≤ -1$.

-   Post-treatment periods are always positive $R ≥ 0$.

For the control group who do not get treated ever, there is no $R$ value for the observations.

## Dynamic Treatment Effects

We can now estimate causal effects for each individual post-treatment time-period $t= 0, 1, 2, \dots$.

```{r}
#| warning: false
#| message: false
#| fig-height: 3.5
#| fig-align: center

library(fixest)
data(base_stagg, package = "fixest")
df <- base_stagg
colnames(df) <- c("unit", "time", "cohort", "rel.time", "treated", "te", "covar", "outcome")
df$treat <- ifelse(df$rel.time >= 0, 1, 0)

dynamic <- feols(
  fml   = outcome ~ i(rel.time, ref = c(-1, -1000)) + covar | unit + time,
  # Do not change -1 in c(). You change -1000 to the value of never-treated for rel.time
  
  data  = df,     # your data
  vcov  = ~unit   # cluster SE by unit
)

library(ggfixest)

plot <- ggiplot(
  # required options
  object   = dynamic,           # input your dynamic model object
  ref.line = FALSE,             # we will add our own line later
  
  # optional aesthetic options
  xlab     = "Relative Time Period (First treatment starts r=0)",  # x-axis label
  ylab     = "Dynamic Effect",        # y-axis label
  main = ""
)

plot +
  # select how many period to display
  xlim(-1, 8) +
  
  # add reference lines for treatment beginning and 0 causal effect
  geom_vline(xintercept = -0.5, linetype = 3) +
  geom_hline(yintercept = 0, linetype = 3, color = "red") +
  theme_bw() +
  theme(
    axis.title.x = element_text(size = 16),  # Increase x-axis title size
    axis.title.y = element_text(size = 16),  # Increase y-axis title size
    axis.text.x = element_text(size = 16),   # Increase x-axis tick label size
    axis.text.y = element_text(size = 16),  # Increase y-axis tick label size
    legend.title = element_text(size = 16),  # Increase legend title size
    legend.text = element_text(size = 14)    # Increase legend text (labels) size
  )

```

This allows us to see how the effect of treatment evolves over time.

-   You can see treatment starts negative in earlier post-treatment periods, before becoming more positive over time.

## Pre-Treatment Effects

We can also calculate estimated effects for pre-treatment periods $t=-1, -2, \dots$. These test the **parallel trends assumption**.

-   If the estimate is not equal to 0, or close to 0, that is evidence the parallel trends assumption is violated.

```{r}
#| warning: false
#| message: false
#| fig-height: 3.5
#| fig-align: center

plot +
  # select how many period to display
  xlim(-8, 0) +
  
  # add reference lines for treatment beginning and 0 causal effect
  geom_vline(xintercept = -0.5, linetype = 3) +
  geom_hline(yintercept = 0, linetype = 3, color = "red") +
  theme_bw() +
  theme(
    axis.title.x = element_text(size = 16),  # Increase x-axis title size
    axis.title.y = element_text(size = 16),  # Increase y-axis title size
    axis.text.x = element_text(size = 16),   # Increase x-axis tick label size
    axis.text.y = element_text(size = 16),  # Increase y-axis tick label size
    legend.title = element_text(size = 16),  # Increase legend title size
    legend.text = element_text(size = 14)    # Increase legend text (labels) size
  )
```

The more pre-treatment estimates being insignificant, the stronger evidence of meeting parallel trends.

## Conditional Parallel Trends

Generalised DiD allows us to **control/condition** for potential variables that cause violations in parallel trends.

-   Let us say we have some variable $X$, that is correlated with $Y$.

-   If the trend of $X$ values over time in the treated group and control group are different, it is likely that the trend in $Y$ between the treated group and control group is different.

-   Thus, we would want to condition on $X$ - i.e. holding $X$ constant, parallel trends is met.

This allows us to apply Generalised DiD to more situations than classical DiD.

## Two-Way Fixed Effects

The estimator for Generalised DiD is the two-way fixed effects (TWFE) estimator. It is a linear regression:

$$
\hat Y_{it} = \underbrace{\hat\alpha_i + \hat\gamma_t}_\text{fixed effects} + D_{it}\hat\tau_\text{ATT} + \mathbf X_{it}^\top \hat{\pmb\beta}
$$

-   $\mathbf X_{it}^\top$ are the values of the variables we are using to condition for parallel trends. This is optional (only use if parallel trends is not met without them).

-   $\hat\alpha_i$ and $\hat\gamma_t$ are unit and time fixed effects.

The estimated $\hat\tau_\text{ATT}$ from the regression is the overall causal estimate for all post-treatment periods. Standard errors should be clustered by units.

## What are Fixed Effects

Unit fixed effects $\hat\alpha_i$ are intercepts in regression, like $\beta_0$. However, each unit $i$ gets its own intercept value.

-   These account for differences between units. Think of them as a control variable for units.

Time fixed effects $\hat\gamma_t$ are the same, but for time periods $t$. These account for differences between time periods.

The only remaining potential confounders is differences in trends over time between units. But if we meet the parallel trends assumption, this is also controlled for.

Thus, all confounders are controlled for, and thus, treatment $D_{it}$ is exogenous. Thus, we can find the causal effect with TWFE.

## TWFE In R

TWFE in R is as follows:

```{r}
#| echo: true
#| eval: false

library(fixest)

model <- feols(
  fml  = Y ~ D + X1 + X2 | unit + time,
  data = df,
  vcov = ~unit
)
summary(model)
```

-   X1 and X2 are confounders. If no confounders, just don't include any.

## TWFE for Dynamic Treatment Effects

We can also use two-way fixed effects to estimate dynamic treatment effects (post-treatment) and pre-treatment estimates.

$$
\hat Y_{it} = \underbrace{\hat\alpha_i + \hat\gamma_t}_\text{fixed effects} + \sum_{r≠-1}I_{itr}G_i\hat\tau_r + \mathbf X_{it}^\top \hat{\pmb\beta}
$$

-   $r$ is the specific relative time period $R$ of the observation in question (see slide 3).

-   $I_{itr}$ is an indicator, that takes value $I=1$ if observation $it$ is in relative time period $R=r$, and $I=0$ otherwise. $G_i$ is a treatment/control group indicator.

-   The $\sum_r I_{itr}G_i$ basically says estimate a causal effect $\hat\tau_r$ for every relative time period $r$, except -1 (the last pre-treatment period).

## Dynamic TWFE In R

```{r}
#| echo: true
#| eval: false

library(fixest)
library(ggfixest)

model <- feols(
  fml  = Y ~ i(R, group, ref = -1) + X | unit + time,
  data = df,
  vcov = ~unit
)
ggiplot(model)
```

-   R is the relative time period variable.

-   group is a binary variable. Individuals that receive treatment at any time (treatment group) get value 1, individuals that never receive treatment (control group) get value 0.
